DCMIP2016: A Review of Non-hydrostatic Dynamical Core Design and Intercomparison of Participating Models

### 2017-10-26 DCMIP2016-Part1-Models-20171008.tex
- Near final production update

 \usepackage{xcolor} \usepackage{ulem} \usepackage{changebar}
 \newcommand\TLSins[1]{\cbstart{}\textcolor{ins}{\uline{#1}}\cbend{}}
 \newcommand\TLSdel[1]{\cbdelete{}\textcolor{del}{\sout{#1}}}

@@

\Author[4]{Joseph}{Klemp}
\Author[4]{Sang-Hun}{Park}
\Author[4]{William}{Skamarock}

\Author[14]{Hiroaki}{Miura}
\Author[14]{Tomoki}{Ohno}
\Author[15]{Ryuji}{Yoshida}

\Author[16]{Robert}{Walko}

\Author[17]{Alex}{Reinecke}
\Author[17]{Kevin}{Viner}

@@

\affil[1]{University of California, Davis}
\affil[2]{University of Michigan}
\affil[3]{University of South Wales}
\affil[4]{National Center for Atmospheric Research}
\affil[5]{Stony Brook University}
\affil[6]{University of Colorado, Boulder}
\affil[7]{Colorado State University}
\affil[8]{Institut Pierre-Simon Laplace (IPSL)}
\affil[9]{Geophysical Fluid Dynamics Laboratory (GFDL)}
\affil[10]{European Centre for Medium-Range Weather Forecasts (ECMWF)}
\affil[11]{Environment and Climate Change Canada}
\affil[12]{Max Planck Institute for Meteorology}
\affil[13]{Deutscher Wetterdienst (DWD)}
\affil[14]{University of Tokyo}
\affil[15]{RIKEN}
\affil[16]{University of Miami}
\affil[17]{Naval Research Laboratory}

@@

\begin{abstract}
Atmospheric dynamical cores are a fundamental component of global atmospheric modeling systems, and are responsible for capturing the dynamical behavior of the Earth's atmosphere via numerical integration of the Navi\'er-Stokes equations.  These systems have existed in one form or another for over half of a century, with the earliest discretizations having now evolved into a complex ecosystem of algorithms and computational strategies.  In essence, no two dynamical cores are alike, and their individual successes suggest that no perfect model exists.  To better understand modern dynamical cores, this paper aims to provide a comprehensive review of eleven dynamical cores, drawn from modeling centers and groups that participated in the 2016 Dynamical Core Model Intercomparison Project (DCMIP) workshop and summer school.  This review includes choice of model grid, variable placement, vertical coordinate, prognostic equations, temporal discretization, and the diffusion, stabilization, filters and fixers employed by each system.
\end{abstract}

@@

\begin{table}[p]
\caption{A standard list of symbols used throughout this paper and in the DCMIP.} \label{tab:symbols}
\begin{center}
\begin{tabular}{cl}
\hline Symbol & Description \\ \hline 
$\lambda$ & Longitude (in radians) \\
$\varphi$ & Latitude (in radians) \\
$z$ & Height with respect to mean sea level (set to zero) \\
$s$ & Vertical model coordinate \\
$p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
$\Phi_s$ & Surface geopotential \\
$z_s$ & Surface elevation with respect to mean sea level (set to zero) \\
$u$ & Zonal wind velocity \\
$v$ & Meridional wind velocity \\
$w$ & Vertical wind velocity \\
$\vb{u}$ & 3D wind vector \\
$\vb{u}_h$ & Horizontal wind vector \\
$\vb{v}_h$ & Horizontal wind vector with covariant components \\
$\omega$ & Vertical pressure velocity  \\
$D$ & Divergence of the horizontal wind vector\\
$\zeta$ & Vertical component of relative vorticity\\
$p$ & Pressure (pressure of moist air if $q>0$) \\
$e$ & Internal energy \\
$\rho$ & Total air density \\
$\rho_d$ & Dry air density \\
$\rho_s$ & Pseudo-density \\
$T$ &Temperature \\
$T_v$ & Virtual temperature \\
$\theta$ & Potential temperature \\
$\theta_v$ & Virtual potential temperature \\
$\theta_{il}$ & Ice-liquid potential temperature \\
$\theta_\rho$ & Density potential temperature \\
%$P_{ls}$ & Large-scale precipitation rate \\
$q$ & Specific humidity \\
$q_v$ & Water vapor mixing ratio \\
$q_c$ & Cloud water mixing ratio \\
$q_r$ & Rain water mixing ratio \\
$q_i$ & General tracer mixing ratio \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

This section provide a brief overview of key discretization choices, along with unique features or design specifications from participating dynamical cores.  Further details on these choices can be found in subsequent sections.  In total, simulation results and model descriptions have been submitted from eleven dynamical cores (see Table \ref{tab:Models}).  The prognostic variables employed and horizontal discretizations for these dynamical cores are summarized in Table \ref{tab:ModelsHorizontal}.  The vertical staggering of variables and vertical coordinate choice is summarized in Table \ref{tab:ModelsVertical}.  Standard options for diffusion, stabilization, filters, or fixers along with the temporal discretization for these models is summarized in Table \ref{tab:ModelsTemporalDiffusion}.  A brief description of each participant model follows, focused on the unique features and decisions underlying the model design.

@@

\begin{table}[p]
\caption{Participating modeling centers and associated dynamical cores that have submitted a model description and/or simulation results.} \label{tab:Models}
\begin{center}
\begin{tabular}{cll}
\hline Short Name & Long Name & Modeling Center or Group \\ \hline 
ACME--A & Atmosphere model of the Accelerated Climate Model for Energy & Sandia National Laboratories and \\
& & University of Colorado, Boulder, USA \\
CSU & Colorado State University Model & Colorado State University, USA \\
DYNAMICO & DYNAMICO & Institut Pierre Simon Laplace (IPSL), France \\
FV$^3$ & GFDL Finite-Volume Cubed-Sphere Dynamical Core & Geophysical Fluid Dynamics Laboratory, USA \\
FVM & Finite Volume Module of the Integrated Forecasting System & European Centre for Medium-Range Weather Forecasts \\
GEM & Global Environmental Multiscale model & Environment and Climate Change Canada \\
ICON & Icosahedral Non-hydrostatic model & Max-Planck-Institut f\"ur Meteorologie, Germany \\
MPAS & Model for Prediction Across Scales & National Center for Atmospheric Research, USA \\
NICAM & Non-hydrostatic Icosahedral Atmospheric Model & AORI / JAMSTEC / AICS, Japan \\
OLAM & Ocean Land Atmosphere Model & Duke University / University of Miami, USA \\
TEMPEST & Tempest Non-hydrostatic Atmospheric Model & University of California, Davis, USA \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[p]
\caption{Details on the prognostic variables and horizontal discretization for participating dynamical cores.  Equation set indicates whether a model is hydrostatic (H) or non-hydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).  Only three numerical methods are represented among participating models, namely finite-difference (FD), finite-volume (FV), and spectral-element (SE).  More details on horizontal staggering can be found in section \ref{sec:HorizontalStaggering}.} \label{tab:ModelsHorizontal}
\begin{center}
\begin{tabular}{cccccc}
\hline Short Name & Equation Set & Prognostic Variables & Horizontal Grid & Numerical & Horizontal  \\
& & & & method & staggering \\ \hline
ACME--A & H/NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta$, $\Phi$, $\rho_s q_i$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q_i$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
DYNAMICO & H/NH & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q_i$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\delta \Phi$, $\rho_s q_i$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
FVM & NH (D) & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q_i$ & Octahedral (\S \ref{sec:grid_oct}) & FV & A-grid \\
GEM & NH & $\vb{u}_h$, $w$, $\dot{\zeta}$, $T_v$, $p$, $q_i$ & Yin-Yang (\S \ref{sec:grid_yinyang}) & FD & C-grid  \\
ICON & NH (D) & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q_i$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q_i$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q_i$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
OLAM & NH (D) & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho \theta_{il}$, $\rho q_i$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
TEMPEST & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q_i$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

\begin{table}[p]
\caption{Standard options for diffusion, stabilization, filters, or fixers in participating dynamical cores (detailed in section \ref{sec:DiffusionStabilization}) and temporal discretization (detailed in section \ref{sec:TemporalDiscretizations}).} \label{tab:ModelsTemporalDiffusion}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Standard options for diffusion, & Temporal discretization \\
& stabilization, filters, or fixers & \\ \hline 
ACME--A & 4th-order horizontal hyperviscosity & KGU53 \citep{guerra2016high} \\
CSU & 4th-order horizontal hyperviscosity & 3rd-order Adams-Bashforth (AB3) \\
DYNAMICO & 4th-order horizontal hyperviscosity & ARK232 \citep{giraldo2013implicit} \\
FV$^3$ & Divergence damping, hyperviscosity & Forward-backwards \citep{LR1997QJR} / semi-implicit \\
FVM & Monotonic limiting & Semi-implicit \citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}) \\
GEM & Hyperviscosity & Semi-implicit \citep{Girard2014} (\S \ref{sec:GEM_temporal}) \\
ICON & Divergence damping, Smagorinsky, hyperdiffusion & Predictor-corrector \\
MPAS & Smagorinsky, hyperdiffusion & Split-explicit \citep{klemp2007conservative} \\
NICAM & 3D divergence damping, Smagorinsky, hyperviscosity & Split-explicit \citep{klemp2007conservative} \\
OLAM & Divergence/vorticity damping & 2nd-order Adams-Bashforth, Lax-Wendroff (for tracers) \\
TEMPEST & 4th-order horizontal hyperviscosity & ARS232 \citep{ascher1997implicit} \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

The ACME--A model has much in common with the Community Atmosphere Spectral Element Model (CAM-SE) \citep{dennis2012cam} as both share a common origin in the High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible}. ACME-A employs both a hydrostatic model and an experimental non-hydrostatic compressible shallow-atmosphere model. Both variants are designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts. ACME-A partitions the globe horizontally via an unstructured grid of quadrilateral elements arranged in a cubed-sphere configuration, although unstructured regionally-refined meshes with conforming edges may also be employed. The fluid equations are discretized using dimensional splitting, with a nodal 4th-order spectral element discretization in the horizontal and a vertical floating Lagrangian levels in hybrid terrain-following pressure coordinates $\eta$.
Vertical operators are based on the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy}. All fields are co-located in the horizontal, in the sense that they share the same 4th-order basis functions.

@@

The CSU model uses an optimized geodesic grid to discretize the sphere \citep{heikes1995numerical,heikes2013optimized}, with height as the vertical coordinate. The model is based on the non-hydrostatic Unified System of equations proposed by \cite{arakawa2009unification}, which filters vertically propagating sound waves but allows the Lamb wave and does not require a reference state. The horizontal wind field is determined by predicting the vertical component of the vorticity and the divergence of the horizontal wind, and then solving a pair of two-dimensional Poisson equations for a stream function and velocity potential. Horizontal diffusion is included in the form of a fourth-order Laplacian operator applied on constant height surfaces ($\nabla^4_z$) that acts on the vorticity, divergence, potential temperature, and tracer.  The CSU model supports both third-order and fifth-order upstream-weighted finite-volume advection schemes, with positivity preservation enforced via mass borrowing.
\subsection{DYNAMICO}

DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations. DYNAMICO's design uniquely combines a representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry, \citep{Dubos2015DYNAMICO10}.  It includes a novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the literature \citep{thuburn2009numerical,ringler2010unified}, and (up to the addition of explicit diffusion) leads to an energy-conserving spatial discretization.  It also incorporates a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation).  In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.

@@

The GFDL Finite-Volume Cubed-Sphere Dynamical Core (FV$^3$, or sometimes written FV3) is a fully finite-volume discretization of the fully-compressible non-hydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid with flow-following Lagrangian vertical coordinate \citep{lin2004vertically}. The Lagrangian vertical coordinate deforms so that the flow is constrained to follow the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms (see section \ref{sec:FloatingLagrangian} below). Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in non-hydrostatic applications the monotonicity constraint is used primarily for tracer transport. The discretization is on the C-D grid as described by \citet{LR1997QJR} (also see section \ref{sec:HorizontalStaggering}). Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.  Implicit viscosity is applied through the monotonicity constraint; if non-monotonic advection is used for the momentum and total air mass a weak explicit hyperviscosity is applied for stability and to alleviate numerical noise. Explicit viscosity is applied every acoustic timestep.

The prognostic horizontal winds are stored in the native Gnomonic local coordinate. The non-hydrostatic solver adds a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density. All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, diagnostic vorticity is a 3D cell-mean value.

@@

The Finite-Volume Module (FVM) of the Integrated Forecasting System (IFS) is currently under development at 
ECMWF \citep{smolarkiewiczetalJCP2016,kuehnleinJCP2017,smolarkiewiczetalJCP2017}. FVM solves the 
compressible Euler equations in a geospherical framework \citep{szmelter2010,smolarkiewiczetalJCP2016}. 
A centered two-time-level semi-implicit 
integration scheme is employed with 3D implicit treatment of acoustic, buoyant, and rotational modes 
\citep{smolarkiewiczJCP2014}. The associated 3D Helmholtz problem is solved iteratively using a bespoke 
preconditioned Generalised Conjugate Residual approach. The integration procedure uses the 
non-oscillatory MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) 
advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}. 
The horizontal spatial discretization is fully unstructured finite-volume using the median-dual approach. This is 
combined with a structured-grid finite-difference discretization in the vertical direction. In both the horizontal and 
the vertical discretization all variables are co-located. 
The median-dual finite-volume mesh in the horizontal is developed about the nodes of the octahedral reduced 
Gaussian grid (Section~\ref{subsection:octahedralreducedgaussiangrid}). The octahedral reduced Gaussian 
grid is also employed in the spectral-transform dynamical core of the presently operational IFS at ECMWF, which facilitates 
interoperability of the two formulations. However, FVM is not restricted to this grid and offers capabilities towards 
broad classes of meshes including adaptivity \citep[e.g.][]{szmelter2010,kuehnleinJCP2012}.
%No explicit diffusion or filtering is applied in FVM for DCMIP, unless it is required for particular test cases.

>>>>

@@

\Author[4]{Joseph}{Klemp}
\Author[14]{Sang-Hun}{Park}
\Author[4]{William}{Skamarock}

\Author[15]{Hiroaki}{Miura}
\Author[15]{Tomoki}{Ohno}
\Author[16]{Ryuji}{Yoshida}

\Author[17]{Robert}{Walko}

\Author[18]{Alex}{Reinecke}
\Author[18]{Kevin}{Viner}

@@

\affil[1]{University of California, Davis}
\affil[2]{University of Michigan}
\affil[3]{University of South Wales}
\affil[4]{National Center for Atmospheric Research}
\affil[5]{Stony Brook University}
\affil[6]{University of Colorado, Boulder}
\affil[7]{Colorado State University}
\affil[8]{Institut Pierre-Simon Laplace (IPSL)}
\affil[9]{Geophysical Fluid Dynamics Laboratory (GFDL)}
\affil[10]{European Centre for Medium-Range Weather Forecasts (ECMWF)}
\affil[11]{Environment and Climate Change Canada}
\affil[12]{Max Planck Institute for Meteorology}
\affil[13]{Deutscher Wetterdienst (DWD)}
\affil[14]{Yonsei University}
\affil[15]{University of Tokyo}
\affil[16]{RIKEN AICS / Kobe University}
\affil[17]{University of Miami}
\affil[18]{Naval Research Laboratory}

@@

\begin{abstract}
Atmospheric dynamical cores are a fundamental component of global atmospheric modeling systems, and are responsible for capturing the dynamical behavior of the Earth's atmosphere via numerical integration of the Navi\'er-Stokes equations.  These systems have existed in one form or another for over half of a century, with the earliest discretizations having now evolved into a complex ecosystem of algorithms and computational strategies.  In essence, no two dynamical cores are alike, and their individual successes suggest that no perfect model exists.  To better understand modern dynamical cores, this paper aims to provide a comprehensive review of eleven non-hydrostatic dynamical cores, drawn from modeling centers and groups that participated in the 2016 Dynamical Core Model Intercomparison Project (DCMIP) workshop and summer school.  This review includes choice of model grid, variable placement, vertical coordinate, prognostic equations, temporal discretization, and the diffusion, stabilization, filters and fixers employed by each system.
\end{abstract}

@@

\begin{table}[p]
\caption{A standard list of symbols used throughout this paper and in the DCMIP.} \label{tab:symbols}
\begin{center}
\begin{tabular}{cl}
\hline Symbol & Description \\ \hline 
$\lambda$ & Longitude (in radians) \\
$\varphi$ & Latitude (in radians) \\
$z$ & Height with respect to mean sea level (set to zero) \\
$s$ & Vertical model coordinate \\
$p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
$\Phi$ & Geopotential \\
$\Phi_s$ & Surface geopotential \\
$z_s$ & Surface elevation with respect to mean sea level (set to zero) \\
$u$ & Zonal wind velocity \\
$v$ & Meridional wind velocity \\
$w$ & Vertical wind velocity \\
$\dot{\zeta}$ & GEM vertical coordinate velocity \\
$\vb{u}$ & 3D wind vector \\
$\vb{u}_h$ & Horizontal wind vector \\
$\vb{v}_h$ & Horizontal wind vector with covariant components \\
$\omega$ & Vertical pressure velocity  \\
$D$ & Divergence of the horizontal wind vector\\
$\zeta$ & Vertical component of relative vorticity\\
$p$ & Pressure (pressure of moist air if $q>0$) \\
$e$ & Internal energy \\
$\rho$ & Total air density \\
$\rho_d$ & Dry air density \\
$\rho_s$ & Pseudo-density \\
$T$ &Temperature \\
$T_v$ & Virtual temperature \\
$\theta$ & Potential temperature \\
$\theta_v$ & Virtual potential temperature \\
$\theta_{il}$ & Ice-liquid potential temperature \\
$\theta_\rho$ & Density potential temperature \\
%$P_{ls}$ & Large-scale precipitation rate \\
$q$ & Specific humidity \\
$q_v$ & Water vapor mixing ratio \\
$q_c$ & Cloud water mixing ratio \\
$q_r$ & Rain water mixing ratio \\
$q_i$ & General tracer mixing ratio \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

This section provides a brief overview of key discretization choices, along with unique features or design specifications from participating dynamical cores.  Further details on these choices can be found in subsequent sections.  In total, simulation results and model descriptions have been submitted from eleven dynamical cores (see Table \ref{tab:Models}).  The prognostic variables employed and horizontal discretizations for these dynamical cores are summarized in Table \ref{tab:ModelsHorizontal}.  The vertical staggering of variables and vertical coordinate choice is summarized in Table \ref{tab:ModelsVertical}.  Principal options for diffusion, stabilization, filters, or fixers along with the temporal discretization for these models is summarized in Table \ref{tab:ModelsTemporalDiffusion}.  A brief description of each participant model follows, focused on the unique features and decisions underlying the model design.

@@

\begin{table}[p]
\caption{Participating modeling centers and associated dynamical cores that have submitted a model description and/or simulation results.} \label{tab:Models}
\begin{center}
\begin{tabular}{cll}
\hline Short Name & Long Name & Modeling Center or Group \\ \hline 
ACME--A & Atmosphere model of the Accelerated Climate Model for Energy & Sandia National Laboratories and \\
& & University of Colorado, Boulder, USA \\
CSU & Colorado State University Model & Colorado State University, USA \\
DYNAMICO & DYNAMical core on the ICOsahedron & Institut Pierre Simon Laplace (IPSL), France \\
FV$^3$ & GFDL Finite-Volume Cubed-Sphere Dynamical Core & Geophysical Fluid Dynamics Laboratory, USA \\
FVM & Finite Volume Module of the Integrated Forecasting System & European Centre for Medium-Range Weather Forecasts \\
GEM & Global Environmental Multiscale model & Environment and Climate Change Canada \\
ICON & ICOsahedral Non-hydrostatic model & Max-Planck-Institut f\"ur Meteorologie, Germany \\
MPAS & Model for Prediction Across Scales & National Center for Atmospheric Research, USA \\
NICAM & Non-hydrostatic Icosahedral Atmospheric Model & AORI / JAMSTEC / AICS, Japan \\
OLAM & Ocean Land Atmosphere Model & Duke University / University of Miami, USA \\
TEMPEST & Tempest Non-hydrostatic Atmospheric Model & University of California, Davis, USA \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[p]
\caption{Details on the prognostic variables and horizontal discretization for participating dynamical cores.  Equation set indicates whether a model is hydrostatic (H) or non-hydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).  Only three numerical methods are represented among participating models, namely finite-difference (FD), finite-volume (FV), and spectral-element (SE).  More details on horizontal staggering can be found in section \ref{sec:HorizontalStaggering}.} \label{tab:ModelsHorizontal}
\begin{center}
\begin{tabular}{cccccc}
\hline Short Name & Equation Set & Prognostic Variables & Horizontal Grid & Numerical & Horizontal  \\
& & & & method & staggering \\ \hline
ACME--A & H/NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta$, $\Phi$, $\rho_s q_i$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q_i$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
DYNAMICO & H/NH & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q_i$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q_i$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
FVM & NH (D) & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q_i$ & Octahedral (\S \ref{sec:grid_oct}) & FV & A-grid \\
GEM & NH & $\vb{u}_h$, $w$, $\dot{\zeta}$, $T_v$, $p$, $q_i$ & Yin-Yang (\S \ref{sec:grid_yinyang}) & FD & C-grid  \\
ICON & NH (D) & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q_i$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q_i$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q_i$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
OLAM & NH (D) & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho \theta_{il}$, $\rho q_i$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
TEMPEST & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q_i$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

\begin{table}[p]
\caption{Principal options for diffusion, stabilization, filters, or fixers in participating dynamical cores (detailed in section \ref{sec:DiffusionStabilization}) and temporal discretization (detailed in section \ref{sec:TemporalDiscretizations}).} \label{tab:ModelsTemporalDiffusion}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Principal options for diffusion, & Temporal discretization \\
& stabilization, filters, or fixers & \\ \hline 
ACME--A & 4th-order horizontal hyperviscosity & KGU53 \citep{guerra2016high} \\
CSU & 4th-order horizontal hyperviscosity & 3rd-order Adams-Bashforth (AB3) \\
DYNAMICO & 4th-order horizontal hyperviscosity & ARK232 \citep{giraldo2013implicit} \\
FV$^3$ & Divergence damping, hyperviscosity & Forward-backwards \citep{LR1997QJR} / semi-implicit \\
FVM & Monotonic limiting & Semi-implicit \citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}) \\
GEM & Hyperviscosity & Semi-implicit \citep{Girard2014} (\S \ref{sec:GEM_temporal}) \\
ICON & Divergence damping, Smagorinsky, hyperdiffusion & Predictor-corrector \\
MPAS & Smagorinsky, hyperdiffusion & Split-explicit \citep{klemp2007conservative} \\
NICAM & 3D divergence damping, Smagorinsky, hyperviscosity & Split-explicit \citep{klemp2007conservative} \\
OLAM & Divergence/vorticity damping & 2nd-order Adams-Bashforth, Lax-Wendroff (for tracers) \\
TEMPEST & 4th-order horizontal hyperviscosity & ARS232 \citep{ascher1997implicit} \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

The ACME--A model has much in common with the Community Atmosphere Spectral Element Model (CAM-SE) \citep{dennis2012cam} as both share a common origin in the High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible}. ACME-A employs both a hydrostatic model and an experimental non-hydrostatic compressible shallow-atmosphere model. Both variants are designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts. ACME-A is built upon an unstructured grid of quadrilateral elements arranged in a cubed-sphere configuration, although unstructured regionally-refined meshes with conforming edges may also be employed. The fluid equations are discretized using dimensional splitting, with a nodal 4th-order spectral element discretization in the horizontal and vertical floating Lagrangian levels in hybrid terrain-following pressure coordinates.
Vertical operators are based on the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy}. All fields are co-located in the horizontal, in the sense that they share the same 4th-order basis functions.  Tracer transport is sub-cycled relative to the hydrodynamics, using the spectral element method, with tracer mass as the prognostic variable.

@@

The CSU model is a finite-volume model using an optimized geodesic grid \citep{heikes1995numerical,heikes2013optimized}, with height as the vertical coordinate. The model is based on the non-hydrostatic Unified System of equations proposed by \cite{arakawa2009unification}, which filters vertically propagating sound waves but allows the Lamb wave and does not require a reference state. The horizontal wind field is determined by predicting the vertical component of the vorticity and the divergence of the horizontal wind, and then solving a pair of two-dimensional Poisson equations for a stream function and velocity potential. Horizontal diffusion is included in the form of a fourth-order hyperviscosity operator applied on constant height surfaces ($\nabla^4_z$) that acts on the vorticity, divergence, potential temperature, and tracer.  The CSU model supports both third-order and fifth-order upstream-weighted finite-volume advection schemes, with positivity preservation enforced via mass borrowing.
\subsection{DYNAMICO}

DYNAMICO is a mimetic finite-difference / finite-volume model using a geodesic grid and a floating vertical mass coordinate.  Although originally a hydrostatic model, it has been recently extended to solve the shallow-atmosphere non-hydrostatic Euler equations. DYNAMICO's design uniquely combines a representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry \citep{Dubos2015DYNAMICO10}.  It includes a novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the literature \citep{thuburn2009numerical,ringler2010unified}, and (up to the addition of explicit diffusion) leads to an energy-conserving spatial discretization.  It also incorporates a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation).  Horizontal diffusion is included via a fourth-order hyperviscosity operator.  In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.

@@

The GFDL Finite-Volume Cubed-Sphere Dynamical Core (FV$^3$, or sometimes written FV3) is a finite-volume model that solves the non-hydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid with a floating Lagrangian vertical coordinate. The Lagrangian vertical coordinate deforms so that the flow is constrained to follow the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms (see section \ref{sec:FloatingLagrangian} below). The non-hydrostatic formulation extends the hydrostatic model described in  \cite{lin2004vertically} by adding a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density.  The discretization is on the C-D grid as described by \citet{LR1997QJR} (also see section \ref{sec:HorizontalStaggering}), although the prognostic horizontal winds are stored in the native Gnomonic local coordinate.  All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, diagnostic vorticity is a 3D cell-mean value.  Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in non-hydrostatic applications the monotonicity constraint is used primarily for tracer transport. Since divergence is effectively invisible to the solver, a 2D divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.  Implicit viscosity is applied through the monotonicity constraint; if non-monotonic advection is used for the momentum and total air mass a weak explicit hyperviscosity is applied for stability and to alleviate numerical noise. Explicit viscosity is applied every acoustic timestep.

@@

The Finite-Volume Module (FVM) of the Integrated Forecasting System (IFS) is currently under development at 
ECMWF \citep{smolarkiewiczetalJCP2016,kuehnleinJCP2017,smolarkiewiczetalJCP2017}. FVM solves the non-hydrostatic Euler equations on an octahedral reduced Gaussian grid with a height-based terrain-following vertical coordinate \citep{szmelter2010,smolarkiewiczetalJCP2016}. The horizontal spatial discretization uses the median-dual finite-volume approach, combined with a structured-grid finite-difference method in the vertical. In both the horizontal and 
the vertical discretization all variables are co-located. A centered two-time-level semi-implicit 
integration scheme is employed with 3D implicit treatment of acoustic, buoyant, and rotational modes 
\citep{smolarkiewiczJCP2014}. The associated 3D Helmholtz problem is solved iteratively using a bespoke 
preconditioned Generalised Conjugate Residual approach. The integration procedure uses the 
non-oscillatory finite-volume MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) 
advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}.  The non-oscillatory (i.e.~monotonic) MPDATA 
also provides sufficient dissipation/diffusion to stabilize the model, so no other explicit filtering mechanism is required. 
Note that the octahedral reduced Gaussian 
grid is also employed in the spectral-transform dynamical core of the presently operational IFS at ECMWF, which facilitates 
interoperability of the two formulations. However, FVM is not restricted to this grid and offers capabilities towards 
broad classes of meshes \citep[][]{szmelter2010,kuehnleinJCP2012,deconinck2017atlas}.
